High-level cognition during story listening is reflected in high-order dynamic correlations in neural activity patterns

### 2019-09-05 paper/main.tex
another pass through the text, updated patterns figure

@@

\author{Lucy L. W. Owen$^1$, Thomas H. Chang$^{1,2}$, and\
  Jeremy R. Manning\textsuperscript{$1, \dagger$}\\
  [0.1in]$^1$Department of Psychological and Brain
  Sciences,\\Dartmouth
  College, Hanover, NH\\
  $^3$Amazon.com, Seattle, WA\\
  \textsuperscript{$\dagger$}Address correspondence to
  jeremy.r.manning@dartmouth.edu}

@@

\begin{abstract}
  Our thoughts arise from coordinated patterns of interactions between
  brain structures that change with our ongoing experiences.
  High-order dynamic correlations in brain activity patterns reflect
  different subgraphs of the brain's connectome that display
  homologous lower-level dynamic correlations.  We tested the
  hypothesis that high-level cognition is supported by high-order
  dynamic correlations in brain activity patterns.  We developed an
  approach to estimating high-order dynamic correlations in timeseries
  data, and we applied the approach to neuroimaging data collected as
  human participants either listened to a ten-minute story, listened
  to a temporally scrambled version of the story, or underwent a
  resting state scan.  We trained across-participants pattern
  classifiers to decode (in held-out data) when in the session each
  neural activity snapshot was collected.  We found that classifiers
  trained to decode from high-order dynamic correlations yielded the
  best performance on data collected as participants listened to the
  (unscrambled) story.  By contrast, classifiers trained to decode
  data from scrambled versions of the story or during the resting
  state scan yielded the best performance when they were trained using
  first-order dynamic correlations or non-correlational activity patterns.  We
  suggest that as our thoughts become more complex, they are supported
  by higher-order patterns of dynamic network interactions throughout
  the brain.
\end{abstract}

@@

\section*{Introduction}
A central goal in cognitive neuroscience is to elucidate the
\textit{neural code}: the mapping between (a) mental states or
cognitive representations and (b) neural activity patterns. One means
of testing models of the neural code is to ask how accurately that
model is able to ``translate'' neural activity patterns into known (or
hypothesized) mental states or cognitive
representations~\citep[e.g.,][]{HaxbEtal01, NormEtal06, TongPrat12,
  MitcEtal08a, KamiTong05, NishEtal11, PereEtal18, HuthEtal12,
  HuthEtal16}.  Training decoding models on different types of neural
features (Fig.~\ref{fig:patterns}a) can also help to elucidate which
specific aspects of neural activity patterns are informative about
cognition-- and, by extension, which types of neural activity patterns
might comprise the neural code.  For example, prior work has used
region of interest analyses to estimate the anatomical locations of
specific neural representations~\citep[e.g.,][]{EtzeEtal09}, or to
compare the relative contributions to the neural code of multivariate
activity patterns versus patterns of dynamic correlations between
neural activity patterns~\citep[e.g.,][]{MannEtal18, FongEtal19}.  An
emerging theme in this literature is that cognition is mediated by
complex dynamic interactions between brain
structures~\citep{SporHone06, BassEtal06, Turk13, DemeEtal19,
  SoloEtal19, LuriEtal18, PretEtal17, ZouEtal19, MackEtal17}.

@@

Studies of the neural code to date have primarily focused on
univariate or multivariate neural patterns~\citep[for review
see][]{NormEtal06}, or (more recently) on patterns of dynamic
first-order correlations~\citep[i.e., interactions between pairs of
brain structures;][]{MannEtal18, FongEtal19, LuriEtal18, PretEtal17,
  ZouEtal19, DemeEtal19}.  We wondered what the
future of this line of work might hold.  For example, is the neural
code mediated by higher-order interactions between brain
structures~\citep[e.g., see][]{ReimEtal17}?
Second-order correlations reflect \textit{homologous} patterns of
correlation.  In other words, if the changing patterns of correlations
between two regions, $A$ and $B$, are similar to those between two
other regions, $C$ and $D$, this would be reflected in the
second-order correlations between ($A$--$B$) and ($C$--$D$).  In this
way, second-order correlations identify similarities and differences
between subgraphs of the brain's connectome.  Analogously, third-order
correlations reflect homologies between second-order correlations--
i.e., homologous patterns of homologous interactions between brain
regions.  More generally, higher-order correlations reflect homologies
between patterns of lower-order correlations.  We can then ask: which
``orders'' of interaction are most reflective of high-level cognitive
processes?

@@

We validated our approach using synthetic data where the underlying
correlations were known.  We then applied our framework to a
neuroimaging dataset collected as participants listened to either an
audio recording of a ten-minute story or a temporally scrambled
version of the story, or underwent a resting state
scan~\citep{SimoEtal16}.  We used a subset of the data to train
across-participant classifiers to decode listening times using a blend
of neural features (comprising neural activity patterns, as well as
different orders of dynamic correlations between those patterns that
were inferred using our computational framework).  We found that both
the PCA-based and eigenvector centrality-based approaches yielded
neural patterns that could be used to decode accurately (well above
chance).  Both approaches also yielded the best decoding accuracy for
data collected during (intact) story listening when high-order (PCA:
second-order; eigenvector centrality: fourth-order) dynamic
correlation patterns were included as features.  When we trained
classifiers on the scrambled stories or resting state data, only
lower-order dynamic patterns were informative to the decoders.  Taken
together, our results indicate that high-level cognition is supported
by high-order dynamic patterns of communication between brain
structures.

\section*{Methods}
Our general approach to comprises four general steps
(Fig.~\ref{fig:methods}).  First, we derive a kernel-based approach to
computing dynamic pairwise correlations in a $T$ (timepoints) by $K$
(features) multivariate timeseries, $\mathbf{X}_0$.  This yields a $T$
by $\mathcal{O}(K^2)$ matrix of dynamic correlations, $\mathbf{Y}_1$,
where each row comprises the upper triangle of the correlation matrix
at a single timepoint, reshaped into a row vector (this reshaped
vector is $(\frac{K^2 - K}{2})$-dimensional).  Second, we apply a
dimensionality reduction step to project the matrix of dynamic
correlations back onto a $K$-dimensional space.  This yields a $T$ by
$K$ matrix, $\mathbf{X}_1$, that reflects an approximation of the
dynamic correlations reflected in the original data.  Third, we use
repeated applications of the kernel-based dynamic correlation step to
$\mathbf{X}_n$ and the dimensionality reduction step to the resulting
$\mathbf{Y}_{n+1}$ to estimate high-order dynamic correlations.  Each
application of these steps to a $T$ by $K$ time series $\mathbf{X}_n$
yields a $T$ by $K$ matrix, $\mathbf{X}_{n+1}$, that reflects the
dynamic correlations between the columns of $\mathbf{X}_n$.  In this
way, we refer to $n$ as the \textit{order} of the timeseries, where
$\mathbf{X}_0$ (order 0) denotes the original data and $\mathbf{X}_n$
denotes (approximated) $n^\mathrm{th}$-order dynamic correlations
between the columns of $\mathbf{X}_0$.  Finally, we use a
cross-validation--based decoding approach to evaluate how well
information contained in a given order (or weighted mixture of orders)
may be used to decode relevant cognitive states.  If including a given
$\mathbf{X}_n$ in the feature set yields higher classification
accuracy on held-out data, we interpret this as evidence that the
given cognitive states are reflected in patterns of
$n^\mathrm{th}$-order correlations.  All of the code used to produce
the figures and results in this manuscript, along with links to the
corresponding datasets, may be found at
\href{https://github.com/ContextLab/timecorr-paper}{github.com/ContextLab/timecorr-paper}.
In addition, we have released a Python toolbox for computing dynamic
high-order correlations in timeseries data; our toolbox may be found
at
\href{https://timecorr.readthedocs.io/}{timecorr.readthedocs.io}. \textbf{JRM
  NOTE: CHECK LINK}

@@

\subsection*{Kernel-based approach for computing dynamic correlations}
Given a $T$ by $K$ matrix of observations, $\mathbf{X}$, we can compute the (static)
Pearson's correlation between any pair of columns, $\mathbf{X}(\cdot, i)$ and
$\mathbf{X}(\cdot, j)$ using~\citep{Pear01}:
\begin{align}
  \mathrm{corr}(\mathbf{X}(\cdot, i), \mathbf{X}(\cdot, j)) &=
                                                              \frac{\sum_{t=1}^T
                                                              \left(\bar{\mathbf{X}}(t,
                                                              i)
                                                              -
                                                              \mathbf{X}(\cdot,
                                                              i)\right)
                                                              \left(\mathbf{X}(t,
                                                              j)
                                                              -
                                                              \bar{\mathbf{X}}(\cdot, j)\right)}{\sqrt{\sum_{t=1}^T
                                                              \sigma^2_{\mathbf{X}(\cdot, i)} 
                                                              \sigma^2_{\mathbf{X}(\cdot, j)}}},~\mathrm{where}\\\label{eqn:corr}
  \bar{\mathbf{X}}(\cdot, k) &= \frac{1}{T}\sum_{t=1}^T
                       \mathbf{X}(t, k),~\mathrm{and}\\
  \sigma^2_{\mathbf{X}(\cdot, k)} &= \frac{1}{T}\sum_{t=1}^T \left( \mathbf{X}(t, k) -
                            \bar{\mathbf{X}}(\cdot, k) \right)^2 
\end{align}
We can generalize this formula to compute time-varying correlations by
incorporating a \textit{kernel function} that takes a time $t$ as
input, and returns how much the observed data at each timepoint
$\tau \in \left[ -\infty, \infty \right]$ contributes to the estimated instantaneous
correlation at time $t$~\citep[Fig.~\ref{fig:kernels}; also see][for a
similar approach]{AlleEtal12b}.
\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{figs/kernels}
  \caption{\textbf{Examples of kernel functions.} Each panel
    displays per-timepoint weights at $t = 50$, evaluated for 100
    timepoints ($\tau \in \left[1, ..., 100\right]$).  \textbf{a. Uniform
      kernel.} The weights are timepoint-invariant; observations at
    all timepoints are weighted equally, and do not change as a
    function of $\tau$.  This is a special case of weight function that
    reduces dynamic correlations to static correlations.
    \textbf{b. Dirac $\delta$ kernel.} Only the observation at
    timepoint $t$ is given a non-zero weight (of 1).
    \textbf{c. Gaussian kernels.} Each kernel's weights fall off
    in time according to a Gaussian probability density function
    centered on time $t$.  Weights derived using several different
    example width parameters ($\sigma^2$) are displayed.
    \textbf{d. Laplace kernels.}  Each kernel's weights fall off
    in time according to a Laplace probability density function
    centered on time $t$.  Weights derived using several different
    example width parameters ($b$) are displayed.  \textbf{e. Mexican
      hat (Ricker wavelet) kernels.}  Each kernel's weights fall
    off in time according to a Ricker wavelet centered on time $t$.  This
    function highlights the \textit{contrasts} between local versus
    surrounding activity patterns in estimating dynamic
    correlations. Weights derived using several different example
    width parameters ($\sigma$) are displayed.}
  \label{fig:kernels}
\end{figure}

Given a kernel function $\kappa_t(\cdot)$ for timepoint $t$,
evaluated at timepoints $\tau \in \left[ 1, ..., T \right]$, we
can update the static correlation formula in Equation~\ref{eqn:corr}
to estimate the \textit{instantaneous correlation} at timepoint $t$:
\begin{align}
  \mathrm{timecorr}_{\kappa_t}\left(\mathbf{X}(\cdot, i), \mathbf{X}(\cdot, j)\right) &= \frac{\sum_{\tau=1}^T \left( \mathbf{X}(\tau, i) -
                                       \widetilde{\mathbf{X}}_{\kappa_t}(\cdot,
                                                                                        i) \right)
                                 \left( \mathbf{X}(\tau, j) -
                                        \widetilde{\mathbf{X}}_{\kappa_t}(\cdot,
                                                                                        j)\right)}
              {\sqrt{\sum_{\tau=1}^T
                                              \widetilde{\sigma}_{\kappa_t}^2(\mathbf{X}(\cdot,
                                                                                        i))
                                              \widetilde{\sigma}_{\kappa_t}^2(\mathbf{X}(\cdot, j))}},~\mathrm{where}\label{eqn:timecorr}\\
  \widetilde{\mathbf{X}}_{\kappa_t}(t) &= \sum_{\tau=1}^T
                       \kappa_t(\tau, k)\mathbf{X}(\tau, k),\\
  \widetilde{\sigma}_{\kappa_t}^2(\mathbf{X}(\cdot, k)) &= \sum_{\tau=1}^T
                                                  \left(
                                                  \mathbf{X}(\tau, k) -
                            \widetilde{\mathbf{X}}_{\kappa_t}(t, k) \right)^2.
\end{align}
Here
$\mathrm{timecorr}_{\kappa_t}(\mathbf{X}(\cdot, i), \mathbf{X}(\cdot,
j))$ reflects the correlation at time $t$ between columns $i$ and $j$
of $\mathbf{X}$, estimated using the kernel $\kappa_t$.  We evaluate
Equation~\ref{eqn:timecorr} in turn each pair of columns in
$\mathbf{X}$ and for kernels centered on each timepoint in the
timeseries, respectively, to obtain a $T$ by $K$ by $K$ timeseries of
dynamic correlations, $\mathbf{Y}$.  For convenience, we then reshape
the upper triangles and diagonals of each timepoint's correlation matrix into a row
vector to obtain an equivalent $T$ by $\frac{K^2 - K}{2} + K$ matrix.

\subsubsection*{Dynamic inter-subject functional connectivity (DISFC)}
Equation~\ref{eqn:timecorr} provides a means of taking a single
observation matrix, $\mathbf{X}_n$ and estimating the dynamic
correlations from moment to moment, $\mathbf{Y}_{n+1}$.  Suppose that
one has access to a set of multiple observation matrices that reflect
the same phenomenon.  For example, one might collect neuroimaging data
from several experimental participants, as each participant performs
the same task (or sequence of tasks).  Let $\mathbf{X}_n^1$,
$\mathbf{X}_n^2$, ..., $\mathbf{X}_n^P$ reflect the $T$ by $K$
observation matrices ($n = 0$) or reduced correlation matrices
($n > 0$) for each of $P$ participants in an experiment.  We can use
\textit{inter-subject functional connectivity}~\citep[ISFC;
][]{SimoEtal16} to compute the degree of stimulus-driven correlations
reflected in the multi-participant dataset at a given timepoint $t$
using:
\begin{align}
\bar{\mathbf{C}}(t) = M\left(R\left(\frac{1}{2P} \sum_{p=1}^P
  Z\left(Y_n^p(t)\right)^\mathrm{T} + Z\left(Y_n^p(t)\right)\right)\right),\label{eqn:disfc}
\end{align}
where $M$ extracts and vectorizes the diagonal and upper triangle of a symmetric
matrix, $Z$ is the Fisher $z$-transformation~\citep{Zar10}:
\begin{align}
Z(r) = \frac{\log(1+r) - \log(1-r)}{2}
\end{align}
$R$ is the inverse of $Z$:
\begin{align}
R(z) = \frac{\exp(2z - 1)}{\exp(2z + 1)},
\end{align}
and $\mathbf{Y}_n^p(t)$ denotes the correlation matrix
(Eqn.~\ref{eqn:corr}) between each column of $\mathbf{X}_n^p$ and each
column of the average $\mathbf{X}_n$ from all \textit{other}
participants, $\bar{\mathbf{X}}_n^{ \backslash p}$:
\begin{align}
  \bar{\mathbf{X}}_n^{ \backslash p} = R\left(\frac{1}{P-1}\sum_{q \in
  \backslash p} Z\left( \mathbf{X}_n^q \right) \right),
\end{align}
where $\backslash p$ denotes the set of all participants other than
participant $p$. In this way, the $T$ by $\frac{K^2 - K}{2} + K$ DISFC
matrix $\bar{\mathbf{C}}$ provides a time-varying extension of the ISFC
approach developed by \cite{SimoEtal16}.

@@

Given a $T$ by $\frac{K^2 - K}{2} + K$ matrix of dynamic correlations,
$\mathbf{Y}_n$, we propose two general approaches to computing a $T$
by $K$ low-dimensional representation of these correlations,
$\mathbf{X}_n$.  The first approach uses dimensionality reduction
algorithms to project $\mathbf{Y}_n$ onto a $K$-dimensional space.
The second approach uses graph measures to characterize the
relative positions of each feature ($k \in \left[1, ..., K \right]$)
in the network defined by the correlation matrix at each timepoint.

@@

The modern toolkit of dimensionality reduction algorithms include
Principal Components Analysis~\citep[PCA; ][]{Pear01}, Probabilistic
PCA~\citep[PPCA; ][]{TippBish99}, Exploratory Factor
Analysis~\citep[EFA; ][]{Spea04}, Independent Components
Analysis~\citep[ICA; ][]{JuttHera91, ComoEtal91}, $t$-Stochastic
Neighbor Embedding~\citep[$t$-SNE; ][]{MaatHint08}, Uniform Manifold
Approximation and Projection~\citep[UMAP; ][]{McInHeal18},
non-negative matrix factorization~\citep[NMF; ][]{LeeSeun99},
Topographic Factor Analysis (TFA)~\cite{MannEtal14b}, Hierarchical
Topographic Factor analysis (HTFA)~\cite{MannEtal18}, Topographic
Latent Source Analysis~\cite[TLSA; ]{GersEtal11}, Dictionary
learning~\citep{MairEtal09a, MairEtal09b}, deep
auto-encoders~\citep{HintSala06}, among others.  While complete
characterizations of each of these algorithms is beyond the scope of
the present manuscript, the general intuition driving these approaches
is to compute the $T$ by $M$ matrix, $\mathbf{X}$, that is closest
to the original $T$ by $N$ matrix, $\mathbf{Y}$, where (typically)
$M \ll N$.  The different approaches place different constraints on
what properties $\mathbf{X}$ must satisfy and which aspects of
the data are compared (and how) to characterize the match between
$\mathbf{X}$ and $\mathbf{Y}$.

>>>>

\newcommand{\ttests}{S1}
\newcommand{\intact}{S2}
\newcommand{\para}{S3}
\newcommand{\word}{S4}
\newcommand{\rest}{S5}

@@

\author{Lucy L. W. Owen$^1$, Thomas H. Chang$^{1,2}$, and\
  Jeremy R. Manning\textsuperscript{$1, \dagger$}\\
  [0.1in]$^1$Department of Psychological and Brain
  Sciences,\\Dartmouth
  College, Hanover, NH\\
  $^2$Amazon.com, Seattle, WA\\
  \textsuperscript{$\dagger$}Address correspondence to
  jeremy.r.manning@dartmouth.edu}

@@

\begin{abstract}
  Our thoughts arise from coordinated patterns of interactions between
  brain structures that change with our ongoing experiences.
  High-order dynamic correlations in neural activity patterns reflect
  different subgraphs of the brain's connectome that display
  homologous lower-level dynamic correlations.  We tested the
  hypothesis that high-level cognition is supported by high-order
  dynamic correlations in brain activity patterns.  We developed an
  approach to estimating high-order dynamic correlations in timeseries
  data, and we applied the approach to neuroimaging data collected as
  human participants either listened to a ten-minute story, listened
  to a temporally scrambled version of the story, or underwent a
  resting state scan.  We trained across-participants pattern
  classifiers to decode (in held-out data) when in the session each
  neural activity snapshot was collected.  We found that classifiers
  trained to decode from high-order dynamic correlations yielded the
  best performance on data collected as participants listened to the
  (unscrambled) story.  By contrast, classifiers trained to decode
  data from scrambled versions of the story or during the resting
  state scan yielded the best performance when they were trained using
  first-order dynamic correlations or non-correlational activity patterns.  We
  suggest that as our thoughts become more complex, they are supported
  by higher-order patterns of dynamic network interactions throughout
  the brain.
\end{abstract}

@@

\section*{Introduction}
A central goal in cognitive neuroscience is to elucidate the
\textit{neural code}: the mapping between (a) mental states or
cognitive representations and (b) neural activity patterns. One means
of testing models of the neural code is to ask how accurately that
model is able to ``translate'' neural activity patterns into known (or
hypothesized) mental states or cognitive
representations~\citep[e.g.,][]{HaxbEtal01, NormEtal06, TongPrat12,
  MitcEtal08a, KamiTong05, NishEtal11, PereEtal18, HuthEtal12,
  HuthEtal16}.  Training decoding models on different types of neural
features (Fig.~\ref{fig:patterns}a) can also help to elucidate which
specific aspects of neural activity patterns are informative about
cognition-- and, by extension, which types of neural activity patterns
might comprise the neural code.  For example, prior work has used
region of interest analyses to estimate the anatomical locations of
specific neural representations~\citep[e.g.,][]{EtzeEtal09}, or to
compare the relative contributions to the neural code of multivariate
activity patterns versus patterns of dynamic correlations between
neural activity patterns~\citep[e.g.,][]{MannEtal18, FongEtal19}.  An
emerging theme in this literature is that cognition is mediated by
dynamic interactions between brain structures~\citep{SporHone06,
  BassEtal06, Turk13, DemeEtal19, SoloEtal19, LuriEtal18, PretEtal17,
  ZouEtal19, MackEtal17}.

@@

Studies of the neural code to date have primarily focused on
univariate or multivariate neural patterns~\citep[for review
see][]{NormEtal06}, or (more recently) on patterns of dynamic
first-order correlations~\citep[i.e., interactions between pairs of
brain structures;][]{MannEtal18, FongEtal19, LuriEtal18, PretEtal17,
  ZouEtal19, DemeEtal19}.  We wondered what the future of this line of
work might hold.  For example, is the neural code mediated by
higher-order interactions between brain structures~\citep[e.g.,
see][]{ReimEtal17}?  Second-order correlations reflect
\textit{homologous} patterns of correlation.  In other words, if the
dynamic patterns of correlations between two regions, $A$ and $B$, are
similar to those between two other regions, $C$ and $D$, this would be
reflected in the second-order correlations between ($A$--$B$) and
($C$--$D$).  In this way, second-order correlations identify
similarities and differences between subgraphs of the brain's
connectome.  Analogously, third-order correlations reflect homologies
between second-order correlations-- i.e., homologous patterns of
homologous interactions between brain regions.  More generally,
higher-order correlations reflect homologies between patterns of
lower-order correlations.  We can then ask: which ``orders'' of
interaction are most reflective of high-level cognitive processes?

@@

We validated our approach using synthetic data where the underlying
correlations were known.  We then applied our framework to a
neuroimaging dataset collected as participants listened to either an
audio recording of a ten-minute story or a temporally scrambled
version of the story, or underwent a resting state
scan~\citep{SimoEtal16}.  We used a subset of the data to train
across-participant classifiers to decode listening times using a blend
of neural features (comprising neural activity patterns, as well as
different orders of dynamic correlations between those patterns that
were inferred using our computational framework).  We found that both
the PCA-based and eigenvector centrality-based approaches yielded
neural patterns that could be used to decode accurately (i.e., well above
chance).  Both approaches also yielded the best decoding accuracy for
data collected during (intact) story listening when high-order (PCA:
second-order; eigenvector centrality: fourth-order) dynamic
correlation patterns were included as features.  When we trained
classifiers on the scrambled stories or resting state data, only
lower-order dynamic patterns were informative to the decoders.  Taken
together, our results indicate that high-level cognition is supported
by high-order dynamic patterns of communication between brain
structures.

\section*{Methods}
Our general approach to efficiently estimating high-order dynamic
correlations comprises four general steps (Fig.~\ref{fig:methods}).
First, we derive a kernel-based approach to computing dynamic pairwise
correlations in a $T$ (timepoints) by $K$ (features) multivariate
timeseries, $\mathbf{X}_0$.  This yields a $T$ by $\mathcal{O}(K^2)$
matrix of dynamic correlations, $\mathbf{Y}_1$, where each row
comprises the upper triangle and diagonal of the correlation matrix at a single
timepoint, reshaped into a row vector (this reshaped vector is
$\left( \frac{K^2 - K}{2} + K \right)$-dimensional).  Second, we apply a dimensionality
reduction step to project the matrix of dynamic correlations back onto
a $K$-dimensional space.  This yields a $T$ by $K$ matrix,
$\mathbf{X}_1$, that reflects an approximation of the dynamic
correlations reflected in the original data.  Third, we use repeated
applications of the kernel-based dynamic correlation step to
$\mathbf{X}_n$ and the dimensionality reduction step to the resulting
$\mathbf{Y}_{n+1}$ to estimate high-order dynamic correlations.  Each
application of these steps to a $T$ by $K$ time series $\mathbf{X}_n$
yields a $T$ by $K$ matrix, $\mathbf{X}_{n+1}$, that reflects the
dynamic correlations between the columns of $\mathbf{X}_n$.  In this
way, we refer to $n$ as the \textit{order} of the timeseries, where
$\mathbf{X}_0$ (order 0) denotes the original data and $\mathbf{X}_n$
denotes (approximated) $n^\mathrm{th}$-order dynamic correlations
between the columns of $\mathbf{X}_0$.  Finally, we use a
cross-validation--based decoding approach to evaluate how well
information contained in a given order (or weighted mixture of orders)
may be used to decode relevant cognitive states.  If including a given
$\mathbf{X}_n$ in the feature set yields higher classification
accuracy on held-out data, we interpret this as evidence that the
given cognitive states are reflected in patterns of
$n^\mathrm{th}$-order correlations.  All of the code used to produce
the figures and results in this manuscript, along with links to the
corresponding datasets, may be found at
\href{https://github.com/ContextLab/timecorr-paper}{github.com/ContextLab/timecorr-paper}.
In addition, we have released a Python toolbox for computing dynamic
high-order correlations in timeseries data; our toolbox may be found
at
\href{https://timecorr.readthedocs.io/}{timecorr.readthedocs.io}. \textbf{JRM
  NOTE: CHECK LINK}

@@

\subsection*{Kernel-based approach for computing dynamic correlations}
Given a $T$ by $K$ matrix of observations, $\mathbf{X}$, we can compute the (static)
Pearson's correlation between any pair of columns, $\mathbf{X}(\cdot, i)$ and
$\mathbf{X}(\cdot, j)$ using~\citep{Pear01}:
\begin{align}
  \mathrm{corr}(\mathbf{X}(\cdot, i), \mathbf{X}(\cdot, j)) &=
                                                              \frac{\sum_{t=1}^T
                                                              \left(\mathbf{X}(t,
                                                              i)
                                                              -
                                                              \bar{\mathbf{X}}(\cdot,
                                                              i)\right)
                                                              \left(\mathbf{X}(t,
                                                              j)
                                                              -
                                                              \bar{\mathbf{X}}(\cdot, j)\right)}{\sqrt{\sum_{t=1}^T
                                                              \sigma^2_{\mathbf{X}(\cdot, i)} 
                                                              \sigma^2_{\mathbf{X}(\cdot, j)}}},~\mathrm{where}\\\label{eqn:corr}
  \bar{\mathbf{X}}(\cdot, k) &= \frac{1}{T}\sum_{t=1}^T
                               \mathbf{X}(t, k),~\mathrm{and}\\
  \sigma^2_{\mathbf{X}(\cdot, k)} &= \frac{1}{T}\sum_{t=1}^T \left( \mathbf{X}(t, k) -
                                    \bar{\mathbf{X}}(\cdot, k) \right)^2 
\end{align}
We can generalize this formula to compute time-varying correlations by
incorporating a \textit{kernel function} that takes a time $t$ as
input, and returns how much the observed data at each timepoint
$\tau \in \left[ -\infty, \infty \right]$ contributes to the estimated instantaneous
correlation at time $t$~\citep[Fig.~\ref{fig:kernels}; also see][for a
similar approach]{AlleEtal12b}.
\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{figs/kernels}
  \caption{\textbf{Examples of kernel functions.} Each panel displays
    per-timepoint weights for a kernel centered at $t = 50$, evaluated
    at 100 timepoints ($\tau \in \left[1, ..., 100\right]$).
    \textbf{a. Uniform kernel.} The weights are timepoint-invariant;
    observations at all timepoints are weighted equally, and do not
    change as a function of $\tau$.  This is a special case kernel
    function that reduces dynamic correlations to static correlations.
    \textbf{b. Dirac $\delta$ kernel.} Only the observation at
    timepoint $t$ is given a non-zero weight (of 1).
    \textbf{c. Gaussian kernels.} Each kernel's weights fall off in
    time according to a Gaussian probability density function centered
    on time $t$.  Weights derived using several different example
    width parameters ($\sigma^2$) are displayed.  \textbf{d. Laplace
      kernels.}  Each kernel's weights fall off in time according to a
    Laplace probability density function centered on time $t$.
    Weights derived using several different example width parameters
    ($b$) are displayed.  \textbf{e. Mexican hat (Ricker wavelet)
      kernels.}  Each kernel's weights fall off in time according to a
    Ricker wavelet centered on time $t$.  This function highlights the
    \textit{contrasts} between local versus surrounding activity
    patterns in estimating dynamic correlations. Weights derived using
    several different example width parameters ($\sigma$) are
    displayed.}
  \label{fig:kernels}
\end{figure}

Given a kernel function $\kappa_t(\cdot)$ for timepoint $t$,
evaluated at timepoints $\tau \in \left[ 1, ..., T \right]$, we
can update the static correlation formula in Equation~\ref{eqn:corr}
to estimate the \textit{instantaneous correlation} at timepoint $t$:
\begin{align}
  \mathrm{timecorr}_{\kappa_t}\left(\mathbf{X}(\cdot, i), \mathbf{X}(\cdot, j)\right) &= \frac{\sum_{\tau=1}^T \left( \mathbf{X}(\tau, i) -
                                       \widetilde{\mathbf{X}}_{\kappa_t}(\cdot,
                                                                                        i) \right)
                                 \left( \mathbf{X}(\tau, j) -
                                        \widetilde{\mathbf{X}}_{\kappa_t}(\cdot,
                                                                                        j)\right)}
              {\sqrt{\sum_{\tau=1}^T
                                              \widetilde{\sigma}_{\kappa_t}^2(\mathbf{X}(\cdot,
                                                                                        i))
                                              \widetilde{\sigma}_{\kappa_t}^2(\mathbf{X}(\cdot, j))}},~\mathrm{where}\label{eqn:timecorr}\\
  \widetilde{\mathbf{X}}_{\kappa_t}(t) &= \sum_{\tau=1}^T
                       \kappa_t(\tau)\mathbf{X}(\tau, k),\\
  \widetilde{\sigma}_{\kappa_t}^2(\mathbf{X}(\cdot, k)) &= \sum_{\tau=1}^T
                                                  \left(
                                                  \mathbf{X}(\tau, k) -
                            \widetilde{\mathbf{X}}_{\kappa_t}(t, k) \right)^2.
\end{align}
Here
$\mathrm{timecorr}_{\kappa_t}(\mathbf{X}(\cdot, i), \mathbf{X}(\cdot,
j))$ reflects the correlation at time $t$ between columns $i$ and $j$
of $\mathbf{X}$, estimated using the kernel $\kappa_t$.  We evaluate
Equation~\ref{eqn:timecorr} in turn each pair of columns in
$\mathbf{X}$ and for kernels centered on each timepoint in the
timeseries, respectively, to obtain a $T$ by $K$ by $K$ timeseries of
dynamic correlations, $\mathbf{Y}$.  For convenience, we then reshape
the upper triangles and diagonals of each timepoint's correlation matrix into a row
vector to obtain an equivalent $T$ by $\left( \frac{K^2 - K}{2} + K \right)$ matrix.

\subsubsection*{Dynamic inter-subject functional connectivity (DISFC)}
Equation~\ref{eqn:timecorr} provides a means of taking a single
observation matrix, $\mathbf{X}_n$ and estimating the dynamic
correlations from moment to moment, $\mathbf{Y}_{n+1}$.  Suppose that
one has access to a set of multiple observation matrices that reflect
the same phenomenon.  For example, one might collect neuroimaging data
from several experimental participants, as each participant performs
the same task (or sequence of tasks).  Let $\mathbf{X}_n^1$,
$\mathbf{X}_n^2$, ..., $\mathbf{X}_n^P$ reflect the $T$ by $K$
observation matrices ($n = 0$) or reduced correlation matrices
($n > 0$) for each of $P$ participants in an experiment.  We can use
\textit{inter-subject functional connectivity}~\citep[ISFC;
][]{SimoEtal16} to compute the stimulus-driven correlations reflected
in the multi-participant dataset at a given timepoint $t$ using:
\begin{align}
\bar{\mathbf{C}}(t) = M\left(R\left(\frac{1}{2P} \sum_{p=1}^P
  Z\left(Y_n^p(t)\right)^\mathrm{T} + Z\left(Y_n^p(t)\right)\right)\right),\label{eqn:disfc}
\end{align}
where $M$ extracts and vectorizes the diagonal and upper triangle of a symmetric
matrix, $Z$ is the Fisher $z$-transformation~\citep{Zar10}:
\begin{align}
Z(r) = \frac{\log(1+r) - \log(1-r)}{2}
\end{align}
$R$ is the inverse of $Z$:
\begin{align}
R(z) = \frac{\exp(2z - 1)}{\exp(2z + 1)},
\end{align}
and $\mathbf{Y}_n^p(t)$ denotes the correlation matrix at timepoint $t$
(Eqn.~\ref{eqn:timecorr}) between each column of $\mathbf{X}_n^p$ and each
column of the average $\mathbf{X}_n$ from all \textit{other}
participants, $\bar{\mathbf{X}}_n^{ \backslash p}$:
\begin{align}
  \bar{\mathbf{X}}_n^{ \backslash p} = R\left(\frac{1}{P-1}\sum_{q \in
  \backslash p} Z\left( \mathbf{X}_n^q \right) \right),
\end{align}
where $\backslash p$ denotes the set of all participants other than
participant $p$. In this way, the $T$ by $\left( \frac{K^2 - K}{2} + K \right)$ DISFC
matrix $\bar{\mathbf{C}}$ provides a time-varying extension of the ISFC
approach developed by \cite{SimoEtal16}.

@@

Given a $T$ by $\left( \frac{K^2 - K}{2} + K \right)$ matrix of
dynamic correlations, $\mathbf{Y}_n$, we propose two general
approaches to computing a $T$ by $K$ low-dimensional representation of
those correlations, $\mathbf{X}_n$.  The first approach uses
dimensionality reduction algorithms to project $\mathbf{Y}_n$ onto a
$K$-dimensional space.  The second approach uses graph measures to
characterize the relative positions of each feature
($k \in \left[1, ..., K \right]$) in the network defined by the
correlation matrix at each timepoint.

@@

The modern toolkit of dimensionality reduction algorithms include
Principal Components Analysis~\citep[PCA;][]{Pear01}, Probabilistic
PCA~\citep[PPCA;][]{TippBish99}, Exploratory Factor
Analysis~\citep[EFA;][]{Spea04}, Independent Components
Analysis~\citep[ICA;][]{JuttHera91, ComoEtal91}, $t$-Stochastic
Neighbor Embedding~\citep[$t$-SNE;][]{MaatHint08}, Uniform Manifold
Approximation and Projection~\citep[UMAP;][]{McInHeal18}, non-negative
matrix factorization~\citep[NMF;][]{LeeSeun99}, Topographic Factor
Analysis~\cite[TFA;][]{MannEtal14b}, Hierarchical Topographic Factor
analysis~\cite[HTFA;][]{MannEtal18}, Topographic Latent Source
Analysis~\cite[TLSA;][]{GersEtal11}, Dictionary
learning~\citep{MairEtal09a, MairEtal09b}, deep
auto-encoders~\citep{HintSala06}, among others.  While complete
characterizations of each of these algorithms is beyond the scope of
the present manuscript, the general intuition driving these approaches
is to compute the $T$ by $M$ matrix, $\mathbf{X}$, that is closest to
the original $T$ by $N$ matrix, $\mathbf{Y}$, where (typically)
$M \ll N$.  The different approaches place different constraints on
what properties $\mathbf{X}$ must satisfy and which aspects of the
data are compared (and how) in order to characterize how well
$\mathbf{X}$ approximates $\mathbf{Y}$.
